Modelling Stochastic Time Delay for Regression

### 2021-02-15 documentation/article/article.tex
Improve wording and formulas.

Consider the simple example where the effect of the input is $1$, as shown in the image below. The observed input is given by the red line, the blue line is when the effect actually occurs. The first effect happens one time point after the input. The second effect happens at the same time as the input. A fixed time delay isn't valid in this case because the time shifts differ.

@@

If one were to model the effect using a fixed time delay and OLS, the estimate would only be half the true value because only one of the outputs is aligned. Obviously, this isn't ideal, we want the parameter estimates to be as close to the real values as possible, regardless of any noise in the lag structures. One can mitigate the problem via time aggregations, however in complicated multivariate cases, this is just not feasible.

@@

Our algorithm then attempts to find the maximum likelihood estimate for the following equation:

@@

where $X(t + \mathcal{T})$ is the matrix of time shifted impulses, $X(t)$ is the observed input and 1 is a vector (1,1,1...1).\\

@@

where $\theta_{f}$ and $\theta_{\mathcal{T}}$ represent the parameters of the model $f$ and time shifts $\mathcal{T}$. For simplicity, we assume that the time shift distribution and error distributions are independent.

@@

Before optimisation, the values of each individual time shift ($\tau$) are not known. In addition, the prediction errors $\varepsilon$ can only be calculated if each value of $\tau$ is available (because for each time shift there is a different prediction and hence prediction error). Therefore, our algorithm finds the optimum set of time shifts in an inner optimisation, while iteratively searching for the optimum parameters of $f$ in an outer optimisation. 
Firstly, we define a parametric form and some initial parameters to be estimated for the function $f$. For simplicity, let's take the univariate linear model where the $f(x(t))$ is parameterised by $\beta$, $\sigma_{\varepsilon}$(error standard deviation)  and $\sigma_{\tau}$ (time error standard deviation). First, we initialise some starting values for each of these parameters.

@@

We initialize the optimisation algorithm with all parameters set to the mean of the $\mathcal{T}$ distribution (e.g. zero for the zero mean gaussian). Next, we randomly select a small number of impulses, with the value treated as a hyperparameter. The shifts are then randomly changed to a value drawn for the T distribution (proposal vector), and the likelihood (both time shift and prediction error) for the proposal is calculated. If the proposal likelihood is higher than the current maximum likelihood, we update our initial estimate to the current best guess. This procedure is done in a loop with the best guess of the set of T improving over time.

Lastly, we optimise over the set of parameters $\beta$, $\sigma_{\varepsilon}$ and $\sigma_{\tau}$, iteratively alternating between the time shift optimisation and the parameter optimisation. For the parameter optimisation, typical methods such as gradient descent, genetic algorithms or annealing can be used.  In our implementation we have used the differential evolution algorithm from scipiy.optimize (REF), because of its ability to handle noisy objective functions. In order to improve convergence, we also standardize all input variables to the range 0-1.

@@

\section{Limitations}
As the length of $x(t)$ increases, more impulses are introduced and the size of the decomposed matrix $X(t)$ also increases. At some point, handling $X(t)$ becomes impractical. To account for this, we assume that distant impulses do not affect each other. Concretely, for the $ith$ and $j_th$ impulse (row) in $X(t)$, separated by time $t$, the effect vectors $f(X_i(t + \tau_i))$ and $f(X_j(t + \tau_j))$ are linearly independent.

Therefore, the matrix $X(t)$ is decomposed into a number of segments, where each segment is 
composed of a smaller number of dependent impulses. To select the right split locations, we search the time series for locations with the lowest impulse density. Given that we wish to split $X(t)$ into independent matrices $U_n(t)$ and $V_m(t)$, 
if we choose the split locations such that $\forall_i$ $\epsilon$ $n$ and  $\forall_j$ $\epsilon$ $m$, 
$f(U_i(t + \tau_i))$ is linearly independent from $f(V_j(t + \tau_j))$, 
then segments will also be independent and the procedure unique. 
In practice, it may not be possible to guarantee complete independence of every segment, 
in which case results may vary. After segmentation, each matrix can be optimised independently, 
in parallel, making the inner optimisation procedure tractable for longer length sequences.

Finally, a note on problem constraints. The likelihood estimates are dependent on the inner ($\mathcal{T}$) optimisation procedure. There is no guarantee that the global maximum will be found during this process, particularly for sequences with a high density of impulses.

@@

The true shift distribution $\mathcal{T}$ is given by $\mathcal{T}$ $~ N(0, 2)$. There is one value of $\tau$ for each of the 17 impulses.

>>>>

Consider the simple example in Figure 1 where the effect of the input is $1$. The observed input is given by the red line, the blue line is when the effect actually occurs. The first effect happens one time point after the input. The second effect happens at the same time as the input. A fixed time delay isn't valid in this case because the time shifts differ.

@@

If one were to model the effect using a fixed time delay and OLS, the estimate would only be half the true value because only one of the outputs is aligned. Obviously, this isn't ideal, we want the parameter estimates to be as close to the real values as possible, regardless of any noise in the lag structures. One can mitigate the problem via time aggregations, however in complicated cases with multiple factors, this is just not feasible.

@@

We then model:

@@

where $X(t + \mathcal{T})$ is the matrix of time shifted impulses, $X(t)$ is the observed input, $1$ is the vector $(1,1,1...1)$ and $\epsilon$ is a noise term.\\

@@

where $\theta_{f}$ and $\theta_{\mathcal{T}}$ represent the parameters of the model $f$ and time shifts $\mathcal{T}$ respectively. For simplicity, we assume that the time shift distribution and error distributions are independent.

@@

Before optimisation, the values of each individual time shift ($\tau$) are not known. In addition, the prediction errors $\varepsilon$ can only be calculated if each value of $\tau$ is available (because for each time shift there is a different prediction and hence prediction error). Therefore, our algorithm finds the optimum set of time shifts in an inner optimisation, while iteratively searching for the optimum parameters of $f$ in an outer optimisation loop. 
Firstly, we define a parametric form and some initial parameters to be estimated for the function $f$. For simplicity, let's take the univariate linear model where the $f(x(t))$ is parameterised by $\beta$, $\sigma_{\varepsilon}$(error standard deviation)  and $\sigma_{\tau}$ (time error standard deviation). First, we initialise some starting values for each of these parameters.

@@

We initialize the optimisation algorithm with all parameters set to the mean of the $\mathcal{T}$ distribution (e.g. zero for the zero mean gaussian). Next, we randomly select a small number of impulses, with the value treated as a hyperparameter. The shifts are then randomly changed to a value drawn for the T distribution (proposal vector), and the likelihood (both time shift and prediction error) for the proposal is calculated. If the proposal likelihood is higher than the current maximum likelihood, we update our estimate. This procedure is done in a loop with the best guess of the set of T improving over time.

Lastly, we optimise over the set of parameters $\beta$, $\sigma_{\varepsilon}$ and $\sigma_{\tau}$, iteratively alternating between the time shift optimisation and the parameter optimisation. For the parameter optimisation, typical methods such as gradient descent, genetic algorithms or annealing can be used.  In our implementation we have used the differential evolution algorithm from scipiy.optimize (REF), because of its ability to handle noisy objective functions. In order to improve convergence, we also standardize all input variables to the range (0,1).

@@

\section{Limitations}
As the length of $x(t)$ increases, more impulses are introduced and the size of the decomposed matrix $X(t)$ also increases. At some point, handling $X(t)$ becomes impractical. To account for this, we assume that distant impulses do not affect each other. Concretely, for the $i_{th}$ and $j_{th}$ impulse (row) in $X(t)$, separated by time $t \rightarrow \infty$, the effect vectors $f(X_i(t + \tau_i))$ and $f(X_j(t + \tau_j))$ are linearly independent.

Therefore, the matrix $X(t)$ is decomposed into a number of segments, where each segment is 
composed of a smaller number of dependent impulses. To select the right split locations, we search the time series for locations with the lowest impulse density. Given that we wish to split $X(t)$ into independent matrices $U_n(t)$ and $V_m(t)$, 
if we choose the split locations such that $\forall_i$ $\epsilon$ $n$ and  $\forall_j$ $\epsilon$ $m$, 
$f(U_i(t + \tau_i))$ is linearly independent from $f(V_j(t + \tau_j))$, 
then segments will also be independent and the procedure unique. 
In practice, it may not be possible to guarantee complete independence of every segment, 
in which case results may vary. After segmentation, each matrix can be optimised independently 
in parallel, making the inner optimisation procedure tractable for longer length sequences.

Finally, a note on problem constraints. The likelihood estimates are dependent on the inner ($\mathcal{T}$) optimisation procedure. This procedure is $NP$-$hard$ and there is no gaurantee that the global maximum will be found, particularly for sequences with a high density of impulses.

@@

The true shift distribution $\mathcal{T}$ is given by $\mathcal{T}$ $~$ $N(0, 2)$. There is one value of $\tau$ for each of the 17 impulses.
